---
title: "Data Analysis #2 Version 2 (75 points total)"
author: "Cody, Mark"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

### Instructions

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks", and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in the template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets**. An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question: How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
library(rockchalk) # base R code replaces requirement for this package
library(car)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

##### Section 1: (5 points)

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}
par(mfrow = c(1,2))
hist(mydata$RATIO, main = "Histogram of RATIO", xlab = "RATIO")
qqnorm(mydata$RATIO, main = "QQPlot of RATIO")
qqline(mydata$RATIO)
par(mfrow = c(1,1))
sprintf("Rockchalk skewness is: %f", rockchalk::skewness(mydata$RATIO))
sprintf("Rockchalk kurtosis is: %f", rockchalk::kurtosis(mydata$RATIO))


```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}
mydata$L_RATIO = log10(mydata$RATIO)
par(mfrow = c(1,2))
hist(mydata$L_RATIO, main = "Histogram of L_RATIO", xlab = "L_RATIO (Log10 Transform of RATIO)")
qqnorm(mydata$L_RATIO, main = "QQPlot of L_RATIO")
qqline(mydata$L_RATIO)
par(mfrow = c(1,1))

sprintf("Rockchalk skewness is: %f", rockchalk::skewness(mydata$L_RATIO))
sprintf("Rockchalk kurtosis is: %f", rockchalk::kurtosis(mydata$L_RATIO))

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(L_RATIO ~ CLASS, data = mydata)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: The Log10 transform of RATIO, L_RATIO, exhibits better conformance to a normal distribution. RATIO has less centrality. This histogram has a longer tail, and the qpplot has many points in the upper right quadrant that stray far from the line. L_RATIO shows a more normally shaped distribution in the histogram, has a tighter qqplot, and has both skewness and kurtosis scores that are closer to zero. ***


##### Section 2 (10 points) ###############################

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
e2a1 = aov(L_RATIO ~ CLASS*SEX, data = mydata)
summary(e2a1)
e2a2 = aov(L_RATIO ~ CLASS + SEX, data = mydata)
summary(e2a2)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The interaction term, CLASS:SEX is far less explanatory than SEX and CLASS alone. The results suggests that variability in L_RATIO is most explained by CLASS, then by SEX. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(e2a2)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: The Tukey post-hoc analysis goes beyond the ANOVA by revealing significance within SEX & CLASS, at each of the levels of SEX & CLASS. With regard to CLASS, differences in L_RATIO are most significantly different when compared across non-adjacent levels. For example, A1-A3, A1-A4, and A1-A5 have significant difference, but not A1-A2. With regard to SEX, differences are significant betwen Males and Infants and between Females and Infants, but not between Males and Females. This supports combining Males and Females into a single category of Adults, because there is no statistical advantage to separating them. ***


######  Section 3: (10 points) ##################

(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}
e3a2a = mydata$VOLUME[which(mydata$TYPE == "ADULT")]
e3a2i = mydata$VOLUME[which(mydata$TYPE == "I")]
par(mfrow = c(1,2))
hist(e3a2i, main = "Histogram of VOLUME of Infants", xlab = "VOLUME")
hist(e3a2a, main = "Histogram of VOLUME of Adults", xlab = "VOLUME")
par(mfrow = c(1,1))

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The histograms show a very different distribution across Infants versus Adults. Adults are very normally distributed. Infants are gathered at the low end of the range from 0 to 300. There are a few Infant outliers which may complicate separating them from adults. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}
mydata$L_SHUCK = log10(mydata$SHUCK)
mydata$L_VOLUME = log10(mydata$VOLUME)

xl3 = mydata$L_VOLUME
yl3 = mydata$L_SHUCK

xv3 = mydata$VOLUME
yv3 = mydata$SHUCK

Class = mydata$CLASS
Type = mydata$TYPE

p1 = ggplot(mydata, aes(xv3, yv3)) + 
  geom_point(aes(color = Class, shape = Class)) + 
  ggtitle("Volume and Shuck by Class") +
  xlab("Volume") + ylab('Shuck') 
  
p2 = ggplot(mydata, aes(xl3, yl3)) +
  geom_point(aes(color = Class, shape = Class)) +
  ggtitle("L_Volume and Shuck by Class") +
  xlab("L_Volume") + ylab("Shuck") 

p3 = ggplot(mydata, aes(xv3, yv3)) +
  geom_point(aes(color = Type, shape = Type)) + 
  ggtitle("Volume and Shuck by Type") +
  xlab("Volume") + ylab("Shuck") 

p4 = ggplot(mydata, aes(xl3, yl3)) + 
  geom_point(aes(color = Type, shape = Type)) +
  ggtitle("L_Volume and Shuck by Type") +
  xlab("L_Volume") + ylab("Shuck") 

grid.arrange(
  p1,
  p2,
  p3,
  p4,
  nrow = 2
)



```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: The log transform of volume (L_Volume) reduces the variability by both Class and Type. This is apparent in the slimmer profile of the graphs on the right versus the fan-shaped spread that occurs with Volume on the left. For linear regression, L_Volume is a better choice as residuals will be reduced and the significance of Class and Type easier to detect. The graphs also show clustering of Class and Type. On the right with L_Volume, classes A1 and A2 are mainly lower left and classes A3,4,5 are upper right. Likewise, infants are mainly lower left and adults mainly upper right. ***


######   Section 4: (5 points) ###################################

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))
```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

e4am = lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(e4am)

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: Across classes, the estimates and probabilities are decreasing. As the graphs of L-SHUCK and L_VOLUME above show increasing concentration from the baseline class, A1 Adults, to classes A2,A3,A4, and A5, so also the analysis reflects an analogous numeric trend. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: Type compares favorably to class a predictor. The combination of lower standard error, low t value, low estimate, and low p-value set it appart from the class variables.  ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

###### Section 5: (5 points) #################################

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}
par(mfrow = c(1,2))
hist(e4am$residuals, main = "Histogram of Residuals", xlab = "Residuals")
qqnorm(e4am$residuals, main = "QQPlot of Residuals")
par(mfrow = c(1,1))
sprintf("Rockchalk skewness is: %f", rockchalk::skewness(e4am$residuals))
sprintf("Rockchalk kurtosis is: %f", rockchalk::kurtosis(e4am$residuals))


```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}

r1 = ggplot(mydata, aes(L_VOLUME, e4am$residuals)) +
  geom_point(aes(color = Class, shape = Class)) + 
  ggtitle("Residuals by Class") +
  ylab("Residuals") + xlab('L_VOLUME') 

r2 = ggplot(mydata, aes(L_VOLUME, e4am$residuals)) +
  geom_point(aes(color = Type, shape = Type)) + 
  ggtitle("Residuals by Type") +
  ylab("Residuals") + xlab('L_VOLUME') 

r3 = ggplot(mydata, aes(L_VOLUME, e4am$residuals)) +
  geom_boxplot(aes(color = Class)) +
  ggtitle("Residuals by Class") + ylab("Residuals") +
  xlab("Classes A1-A5")

r4 = ggplot(mydata, aes(L_VOLUME, e4am$residuals)) +
  geom_boxplot(aes(color = Type)) +
  ggtitle("Residuals by Type") + ylab("Residuals") + 
  xlab("Types Infant (I) and Adult")

grid.arrange(
  r1,
  r2,
  r3,
  r4,
  nrow = 2
)

bartlett.test(e4am$residuals ~ mydata$CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The exploration of residuals shows a near normal distribution in the histogram, boxplots centered evenly around zero, and a very random pattern of error in the dot plots. The normalcy of error indicates the model is a good fit, and that Volume is a good discriminator and useful in harvesting decisions. ***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

#### Section 6: (5 points) ########################

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}

e6df = data.frame(Adults = prop.adults, Infants = prop.infants, Volume = volume.value)

e61 = ggplot() + 
  geom_line(data = e6df, aes(x = Volume, y = Adults, color = "blue")) +
  geom_line(data = e6df, aes(x = Volume, y = Infants, color = "red")) + 
  scale_color_discrete(name = "Type", labels = c("Adults", "Infants")) + 
  ggtitle("Proportion of Adults and Infants Protected") +
  ylab("Proportion") + xlab('Volume') +
  geom_hline(yintercept=.5, color = "#555555") +
  geom_vline(xintercept = split.adults, color = "#555555") +
  geom_vline(xintercept = split.infants, color = "#555555") +
  annotate(geom="text", x=split.adults+50, y=.45, label="384.51", color="#555555") +
  annotate(geom="text", x=split.infants+50, y=.45, label="133.82", color="#555555") 
e61

```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The values displayed on the graph represent two scenarios based on volume. 1) 133.82 maximizes the available harvest. Infants over this value are vulnerable, but very few adults are excluded from harvest. 2) 384.51 maximizes protection. Many more infants are protected from harvest, but more adults are also excluded from harvest. ***

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###### Section 7: (10 points)  #######################

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

plot(volume.value, ((1-prop.adults)-(1-prop.infants)),
     main = "Difference in Harvest Proportions",
     xlab = "Volume", ylab = "Difference in Proportions Harvested",
     pch = 20, col = "#555555")


```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}

e7v = volume.value[which.max(smooth.difference)]

plot(volume.value, ((1-prop.adults)-(1-prop.infants)),
     main = "Difference in Harvest Proportions",
     xlab = "Volume", ylab = "Difference in Proportions Harvested",
     pch = 20, col = "#555555")
lines(volume.value, smooth.difference, lty = 1, lwd = 2, col = "orange")
abline(v = e7v, lty = 2, col = "#555555")
legend(600,.5, legend = "Volume = 262.143", lty = 2, col = "#555555")


```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
e7y = (1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
e7x = (1 - prop.infants)[which.max(smooth.difference)]
sprintf("Proportion of Populations Harvested if Cutoff is %f:", e7v)
sprintf("Infants: %f", e7x )
sprintf("Adults: %f", e7y )
```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

######  Section 8: (10 points)  ###################

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
e8a = volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786
e8x = sum(mydata$VOLUME[mydata$TYPE == "I"] > e8a) / total.infants
e8y = sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > e8a) / total.adults
sprintf("Proportion of Populations Harvested if Cutoff is %f:", e8a)
sprintf("Infants: %f", e8x )
sprintf("Adults: %f", e8y )

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
e8b = volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391
sprintf("Proportion of Populations Harvested if Cutoff is %f:", e8b)
e9x =  sum(mydata$VOLUME[mydata$TYPE == "I"] > e8b) / total.infants
e9y = sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > e8b) / total.adults
sprintf("Infants: %f",e9x )
sprintf("Adults: %f", e9y )

```


##### Section 9: (5 points) ###########

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}
x9 = 1-prop.infants
y9 = 1-prop.adults

plot(x9, y9, 
     main = "ROC curve of adult and infant harvest proportions",
     xlab = "Infant harvest proportion",
     ylab = "Adult harvest proportion",
     pch = 20, col = "#333333"
)
points(e7x, e7y, pch = 21, col = "orange", lwd = 8)
points(e8x, e8y, pch = 21, col = "green", lwd = 8)
points(e9x, e9y, pch = 21, col = "blue", lwd = 8)
legend(.6,.6, legend = c("max difference", "zero A1 infants", "equal harvest"), 
       col = c("orange", "green", "blue"),
       pch = 15, lwd = 3)
abline(coef = c(0,1), lty = 2)

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
auc(x9,y9)

```


##### Section 10: (10 points) ###################

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	
yd7 = sum(mydata$VOLUME > e7v) / nrow(mydata) #% of population harvested under smooth difference rule
yd8 = sum(mydata$VOLUME > e8a) / nrow(mydata) #% of population harvested under no A1 infants rule
yd9 = sum(mydata$VOLUME > e8b) / nrow(mydata) #% of population harvested under equal harvest rule

rownames = c("Max Difference", "Zero A1 Infants", "Equal Harvest")
colnames = c("Volume","TPR", "FPR", "Yield")

e10 = matrix(c(e7v, e8a, e8b,  e7y, e8y, e9y, e7x, e8x, e9x, yd7, yd8, yd9), 
             nrow = 3, byrow = FALSE,
             dimnames = list(rownames, colnames))
e10

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: The greatest yield (~67.5%) comes from the Zero A1 Infants rule; while it maximizes the percentage of adults harvested (True Positive Result - TPR) at ~82.5%, it also eliminates almost a third (~28.7%) of the infant population. The Equal Harvest option equally distributes the error. It has the second highest yield (~62.4%) and reduces the error rate (False Positive Result - FPR) to ~21.7%. Finally, the Max Difference Rule minimizes infant harvest (~17.6), maintains an adult harvest of ~74.1% and yield of ~58.3%.   ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?  4)  What suggestions would you have for planning future abalone studies of this type? 

***Answer: The greatest conservation is achieved by adopting the Max Difference rule, which considers the volume at which the greatest difference is achieved between adults and infants harvested. Because it minimizes the harvest of infants, it guards most against risk of other unknown or unaccounted for mortality events in the population, such as natural predation which may not discriminate. To implement the rule, harvesters must be able to successfully judge or measure volume, and to enforce,  monitoring and verification routines must be in place. Future studies of abalone may expand the list of characteristics observed in view of finding a discriminating variable that is easier to judge by harvesters than volume.   ***